---
title: "One Chart to Rule Them All?"
subtitle: "Can the I Prime Chart Replace Commonly Used Shewhart Control Charts? <hr>"
author: "Anhøj & Mohammed"
date: "`r Sys.Date()` WORK IN PROGRESS"
output: 
  bookdown::html_document2:
  # html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE,
                      cache     = TRUE,
                      message   = FALSE,
                      fig.asp   = 0.5,
                      fig.width = 9,
                      dev       = 'svg')

library(qicharts2)
# library(patchwork)
library(tidyverse)

bac <- read_csv('data/bacteremia.csv', comment = '#')
hba1c <- read_csv('data/diabetes_hba1c.csv', comment = '#')

# Function to plot control limits from the I' chart on top of Shewhart chart
compplot <- function(x, y, n = NULL, chart, ...) {
  p1 <- qic(x, y, n, chart = chart, ...)
  p2 <- qic(x, y, n, chart = 'ip', ...)
  
  p1 +
    geom_line(aes(y = lcl), data = p2$data, colour = 'tomato') +
    geom_line(aes(y = ucl), data = p2$data, colour = 'tomato')
}
```

----

## Abstract

...

## Introduction

Our aim in this paper is to introduce the I prime chart to healthcare practitioners and highlight its strengths and weakness.

* Basic control chart theory: 
    - common vs special cause variation; 
    - compare the between subgroup variation with the within subgroup variation (sigma)
    - signal special causes when the between subgroup variation exceeds expected (3 sigma) (think ANOVA for time series data).

* Different control charts (to calculate sigma) for different data types.

* Prime charts for count data with large denominators -- when theoretical sigma underestimates the actual within subgroup variation.

* The I chart as a "Swiss army knife" for all data types -- sigma estimated using moving ranges (AMR) reflecting the observed variation in data rather than the theoretical variation based on assumptions about the distribution.

* Limitations of the I chart (does not take subgroup size into account).

* The case for the I prime chart.

Research questions:

* Does the I prime chart match the original I chart for measurement data without denominator?

* Does the I chart adjust appropriately for varying subgroup sizes in measurement data with denominator?

* Is the I prime chart a useful substitute for other Shewhart control charts?

----

Statistical Process Control (SPC) is a widely adopted methodology in healthcare to support quality improvement initiatives. A critical technical challenge in applying SPC is the selection of appropriate control charts, as the choice depends on the type of data being analysed. For continuous (or measurement) data collected one at a time -- such as a patient’s daily blood pressure -- the I chart (individuals chart) is the most suitable option. Conversely, for attribute (or count) data -- such as the number of patients admitted to a hospital -- the standard suite of SPC charts, including the P, U, and C charts, are typically recommended.

However, a point of contention arises when dealing with count data that is expressed as a ratio (e.g., percentage of mortality after surgery). Some practitioners advocate for the use of the I chart in such cases, despite its limitations. The I chart is designed to account for variation between subgroups and does consider variation within subgroups, which is a key feature of the standard attribute charts. This limitation makes the I chart a controversial choice for ratio-based count data, as real-world processes often exhibit within-subgroup and between-subgroup variation, both of which should be accommodated for accurate analysis.

Recent advancements have extended the capabilities of the I chart to handle both measurement data and count data with denominators. This improved I chart (which we call the I prime or I\' chart) now offers a more robust solution for analysing processes with inherent variability, making it a valuable tool for healthcare quality improvement efforts. This paper provides a tutorial review of this enhanced I chart, highlighting its applications and advantages in healthcare settings.

----

...

## Materials and methods

To demonstrate the use of I prime charts on aggregated measurement data with varying subgroup sizes (denominators) we used data from a regional outpatient clinic for children with diabetes. For privacy purposes, data were aggregated in advance including the monthly number of children who visited the clinic and their average glycated haemoglobin (HbA1c, a measure of long term blood glucose levels). Date are presented in Table \@ref(tab:tabhba1c).

To construct the control charts, we used the R package `qicharts2`, which contains functions for Shewhart control charts in addition to functions for prime charts for proportions and rates with large denominators and for the I prime chart.

To demonstrate the use of I prime charts on count data we used data on bacteremia cases (Table \@ref(tab:tabbac)).

Data sets and R code are available from the GitHub repository (https://github.com/anhoej/i_prime_chart).

Additionally, we constructed a large number of I prime charts using both random data from normal, binomial, and poisson distributions and several other data sets from clinical practice. See the appendix for examples of additional plots.

## Results

### I prime chart for measurement data with denominator

Figure \@ref(fig:ichart) is an I chart of the monthly HbA1c averages from Table \@ref(tab:tabhba1c). Notice the data point above the upper control limit in April 2020 suggesting a special cause. However, when plotting aggregated measurement data, the I chart does not account for variations in subgroup size.

```{r ichart, fig.cap='I chart of average HbA1c without denominator'}
qic(month, avg_hba1c, 
    data = hba1c, 
    chart = 'i',
    title = NULL,
    ylab = 'mmol / mol',
    xlab = 'Month')
```

The I prime chart in Figure \@ref(fig:ipchart) takes the subgroup size (number of patients) into account and adjusts the control limits correspondingly. April 2020 was the first month of lockdown during Covid-19 in Denmark, and the number of patients (the denominator) seen during this month was significantly lower than usual, which allowed for larger than usual common cause variation in measurements and wider control limits this month. So, when the subgroup size is taken into account, the apparent special cause in Figure \@ref(fig:ichart) is actually within the limits of the expected common cause variation. Also, notice that the centre lines are a bit different (60.6 vs 60.3), because the I prime chart uses the weighted rather than the unweighted mean of the measurements.

```{r ipchart, fig.cap='I prime chart of average HbA1c with denominator'}
qic(month, avg_hba1c * n, n, 
    data = hba1c, 
    chart = 'ip',
    title = NULL,
    ylab = 'mmol / mol',
    xlab = 'Month')
```

### I prime chart vs P and U charts for count data

In the following plots, the control limits from the original Shewhart charts are shown as a grey background with the limits from the I prime chart algorithm superimposed as red lines.

Figure \@ref(fig:pchart) is a P control chart of the percentage of patients with bacteremia from Table \@ref(tab:tabbac) who died within 30 days after infection. As seen, the I prime limits are (in this case) a bit wider. However, the difference is, for practical purposes, negligible.

```{r pchart, fig.cap='P chart of proportion patients who died of bacteremia.<br>Grey background: control limits from original chart. Red lines: control limits from I prime chart'}
compplot(bac$month, bac$deaths, bac$patients, 
         chart = 'p',
         title = NULL,
         xlab = 'Month')
```

This observation aligns with our experience testing various datasets comparing Shewhart control charts (Xbar, P, C, and U) with I prime limits (see Appendix for examples). While the I prime limits are sometimes slightly wider or narrower than the original limits, the differences are rarely significant enough to affect the analysis conclusions.

Figure \@ref(fig:ppchart) plots the same data as Figure \@ref(fig:pchart) but the original control limits have been calculated using the P prime chart.

```{r ppchart, fig.cap='P prime chart of proportion patients who died of bacteremia.'}
compplot(bac$month, bac$deaths, bac$patients, 
         chart = 'pp',
         title = NULL,
         xlab = 'Month')
```

As seen (and expected from theory), the I prime limits match the P prime limits. This observation has been confirmed with other datasets and with the U and U prime chart of infection rates as demonstrated below (figures \@ref(fig:uchart) and \@ref(fig:upchart).

```{r uchart, fig.cap='U chart of infection rates'}
compplot(bac$month, bac$ha_infections, bac$risk_days, 
         chart = 'u', 
         multiply = 10000,
         title = NULL,
         ylab = 'Infections per 10,000 risk days',
         xlab = 'Month')
```

```{r upchart, fig.cap='U prime chart of infection rates'}
compplot(bac$month, bac$ha_infections, bac$risk_days,
         chart = 'up',
         multiply = 10000,
         title = NULL,
         ylab = 'Infections per 10,000 risk days',
         xlab = 'Month')
```

## Discussion and conclusion

* Summary of findings:

    - The I prime charts is an exact match for the original I chart.

    - For measurement and count data, the I prime charts adjusts the control limits and the centre lines to the size of the denominator thus reflecting any variability in subgroup size.

    - The I prime chart is a very close match for U and P prime charts

    - The I prime chart is a reasonable match for Xbar, C, U, and P charts.

* Strengths, limitations, etc.

* Discussion:

    - Are P and U prime charts always the best choice for count data (even with small denominators)? If so, I prime charts may be used for all types of count and measurement data.

    - NB: no "true" control limits exists (except for synthetic data with known parameters), only "useful" limits exists. Shewhart quote about "highbrow statistical theorems".

* Keen to learn what others think of it ...?

* NB: Say something about run charts!

* Conclusion:

    - ...
    

## Data sets

### Diabetes HbA1c data

* 43 observations of 3 variables:
    - month (date): month of measurements
    - avg_hba1c (numeric): average of HbA1c measurements
    - n (integer): number of patients


```{r tabhba1c, tab.cap='Diabetes HbA1c data set.'}
knitr::kable(hba1c)
```

### Bacteremia data

* 24 observations of 5 variables:
    - month (date): month of infection
    - ha_infections (numeric): number of hospital acquired infections
    - risk_days (numeric): number of patient days without infection
    - deaths (numeric): 30-day mortality after all-cause (community + hospital) infection
    - patients (numeric): number of patients with all-cause infection

```{r tabbac, tab.cap='Bacteremia data set.'}
knitr::kable(bac)
```

## Appendix: Supplementary plots

### I and I prime charts from individual random normal measurements

```{r richart, fig.cap='I chart from 24 random data from a normal distribution without denominator.'}
set.seed(1)                   # fixate random number generator
x <- 1:24                     # subgroups
y <- rnorm(24, 20, 3)         # numerators
n <- runif(24, 80, 120)       # denominators

compplot(x, y, chart = 'i', title = NULL)
```

```{r ripchart, fig.cap='I prime chart from 24 random, normal data with denominator.'}
compplot(x, y * n, n, chart = 'i', title = NULL)
```

### Xbar charts from multiple random normal measurements

```{r rxbarchart1, fig.cap='Xbar chart from 24 subgroups of 10 to 20 random normal data. I prime limits a bit wider than Xbar limits.'}
set.seed(1)
x <- rep(1:24, round(runif(24, 10, 20)))
y <- rnorm(length(x), 9)
compplot(x, y, chart = 'xbar', title = NULL)
```

```{r rxbarchart2, fig.cap='Xbar chart from 24 subgroups of 10 to 20 random normal data. I prime limits close to Xbar limits.'}
set.seed(6)
x <- rep(1:24, round(runif(24, 10, 20)))
y <- rnorm(length(x), 9)
compplot(x, y, chart = 'xbar', title = NULL)
```

```{r rxbarchart3, fig.cap='Xbar chart from 24 subgroups of 10 to 20 random normal data. I prime limits a bit wider than Xbar limits.'}
set.seed(8)
x <- rep(1:24, round(runif(24, 10, 20)))
y <- rnorm(length(x), 9)
compplot(x, y, chart = 'xbar', title = NULL)
```

### P and P prime charts from random binomial data

```{r rppchart, fig.cap='P chart of 24 random data from a binomial distribution.'}
set.seed(1)
x <- 1:24
n <- round(runif(24, 100, 120))
y <- rbinom(24, n, 0.1)
compplot(x, y, n, 'p', title = NULL)
```

```{r rpchart, fig.cap='P prime chart of 24 random data from a binomial distribution.'}
compplot(x, y, n, 'pp', title = NULL)
```

### U and U prime charts from random poisson data

```{r ruchart, fig.cap='U chart from 24 random data from a poisson distribution'}
set.seed(1)
x <- 1:24
n <- (runif(24, 90, 110))
y <- rpois(24, 25)
compplot(x, y, n, chart = 'u', title = NULL)
```

```{r rupchart, fig.cap='U prime chart from 24 random data from a poisson distribution.'}
compplot(x, y, n, chart = 'up', title = NULL)
```
