---
title: "Introducing the I'-chart: an improved individuals chart for quality improvement and control"
# author: "Jacob Anhøj, Wayne Taylor & Mohammed Amin Mohammed"
date: "`r Sys.Date()`"
output:
  bookdown::word_document2:
      reference_docx: "paper_template.docx"
      number_sections: false
    # self_contained: true
    # toc: true
    # toc_float: true
    # css: style.css
bibliography: references.bib
csl: bmj-quality-and-safety.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE,
                      # cache     = TRUE,
                      message   = FALSE,
                      fig.asp   = 0.5,
                      fig.width = 9)

library(qicharts2)
library(tidyverse)

bac   <- read_csv('data/bacteremia.csv', comment = '#')
hba1c <- read_csv('data/diabetes_hba1c.csv', comment = '#')

# Function to plot control limits from the I'-chart on top of Shewhart chart
compplot <- function(x, y, n = NULL, chart, ...) {
  p1 <- qic(x, y, n, chart = chart, ...)
  p2 <- qic(x, y, n, chart = 'ip', ...)
  
  p1 +
    geom_line(aes(y = lcl),
              data     = p2$data,
              linetype = 'dashed',
              colour   = 'tomato') +
    geom_line(aes(y = ucl),
              data     = p2$data,
              linetype = 'dashed',
              colour   = 'tomato')
}
```


## Supplementary materials: Plots

### I'-chart for measurement data

```{r ichart, fig.cap='(ref:ichart)'}
qic(month, avg_hba1c,
    data = hba1c,
    chart = 'i',
    title = NULL,
    ylab = 'mmol / mol',
    xlab = 'Month')
```

```{r ipchart, fig.cap='(ref:ipchart)'}
qic(month, avg_hba1c * n, n,
    data = hba1c,
    chart = 'ip',
    title = NULL,
    ylab = 'mmol / mol',
    xlab = 'Month')
```

```{r pchart, fig.cap='(ref:pchart)'}
compplot(bac$month, bac$deaths, bac$patients,
         chart = 'p',
         title = NULL,
         xlab = 'Month')
```

```{r ppchart, fig.cap='(ref:ppchart)'}
compplot(bac$month, bac$deaths, bac$patients,
         chart = 'pp',
         title = NULL,
         xlab = 'Month')
```

```{r upchart, fig.cap='(ref:upchart)'}
compplot(bac$month, bac$ha_infections, bac$risk_days,
         chart = 'up',
         multiply = 10000,
         title = NULL,
         ylab = 'Infections per 10,000 risk days',
         xlab = 'Month')
```

### I- and I'-charts from individual random normal measurements {-}

```{r richart, fig.cap='(ref:richart)'}
set.seed(1)                   # fix random number generator
x <- 1:24                     # subgroups
y <- rnorm(24, 20, 3)         # numerators
n <- runif(24, 80, 120)       # denominators

p1 <- compplot(x, y, chart = 'i', title = NULL)
p2 <- compplot(x, y * n, n, chart = 'i', title = NULL)

ylim <- range(layer_scales(p1)$y$range$range,
              layer_scales(p2)$y$range$range)
ylim <- scale_y_continuous(limits = ylim)
p1 + ylim
```

```{r ripchart, fig.cap='(ref:ripchart)'}
p2 + ylim
```

### Xbar-charts from multiple random normal measurements {-}

```{r rxbarchart1, fig.cap='(ref:rxbarchart1)'}
set.seed(1)
x <- rep(1:24, round(runif(24, 10, 20)))
y <- rnorm(length(x), 9)
p1 <- compplot(x, y, chart = 'xbar', title = NULL)

set.seed(6)
x <- rep(1:24, round(runif(24, 10, 20)))
y <- rnorm(length(x), 9)
p2 <- compplot(x, y, chart = 'xbar', title = NULL)

set.seed(8)
x <- rep(1:24, round(runif(24, 10, 20)))
y <- rnorm(length(x), 9)
p3 <- compplot(x, y, chart = 'xbar', title = NULL)

ylim <- range(layer_scales(p1)$y$range$range,
              layer_scales(p2)$y$range$range,
              layer_scales(p3)$y$range$range)
ylim <- scale_y_continuous(limits = ylim)

p1 + ylim

```

```{r rxbarchart2, fig.cap='(ref:rxbarchart2)'}

p2 + ylim
```

```{r rxbarchart3, fig.cap='(ref:rxbarchart3)'}

p3 + ylim
```

### P- and P'-charts from random binomial data {-}

```{r rpchart, fig.cap='(ref:rpchart)'}
set.seed(1)
x <- 1:24
n <- round(runif(24, 100, 120))
y <- rbinom(24, n, 0.1)

p1 <- compplot(x, y, n, 'p', title = NULL)
p2 <- compplot(x, y, n, 'pp', title = NULL)


ylim <- range(layer_scales(p1)$y$range$range,
              layer_scales(p2)$y$range$range)
ylim <- scale_y_continuous(limits = ylim)

p1 + ylim
```

```{r rppchart, fig.cap='(ref:rppchart)'}
p2 + ylim
```

### U- and U'-charts from random poisson data {-}

```{r ruchart, fig.cap='(ref:ruchart)'}
set.seed(2)
x <- 1:24
n <- (runif(24, 90, 110))
y <- rpois(24, 25)

p1 <- compplot(x, y, n, chart = 'u', title = NULL)
p2 <- compplot(x, y, n, chart = 'up', title = NULL)

ylim <- range(layer_scales(p1)$y$range$range,
              layer_scales(p2)$y$range$range)
ylim <- scale_y_continuous(limits = ylim)

p1 + ylim
```

```{r rupchart, fig.cap='(ref:rupchart)'}

p2 + ylim
```

(ref:ichart) I-chart of average HbA1c without denominator. The grey region represents the region between the control limits.

(ref:ipchart) I'-chart of average HbA1c with denominator.

(ref:pchart) P-chart of proportion patients who died of bacteremia. Grey background: control limits from P-chart. Dashed lines: control limits from I'-chart.

(ref:ppchart) P'-chart of proportion patients who died of bacteremia. Grey background: control limits from P'-chart.  Dashed lines: control limits from I'-chart.

(ref:upchart) U'-chart of infection rates. Grey background: control limits from U'-chart.  Dashed lines: control limits from I'-chart.

(ref:richart) I-chart from 24 random data from a normal distribution (mean = 20, SD = 3, seed = 1) without denominator. Grey background: control limits from I-chart.  Dashed lines: control limits from I'-chart.

(ref:ripchart) I'-chart from 24 random, normal data (mean = 20, SD = 3, seed = 1) with denominators ranging from 80 to 120. Grey background: control limits from I-chart.  Dashed lines: control limits from I'-chart.

(ref:rxbarchart1) Xbar-chart from 24 subgroups of 10 to 20 random normal data (mean = 9, SD = 1, seed = 1). I' limits a bit wider than Xbar limits. Grey background: control limits from Xbar-chart.  Dashed lines: control limits from I'-chart.

(ref:rxbarchart2) Xbar-chart from 24 subgroups of 10 to 20 random normal data (mean = 9, SD = 1, seed = 6). I' limits close to Xbar limits. Grey background: control limits from Xbar-chart.  Dashed lines: control limits from I'-chart.

(ref:rxbarchart3) Xbar-chart from 24 subgroups of 10 to 20 random normal data (mean = 9, SD = 1, seed = 8). I' limits a bit tigther than Xbar limits. Grey background: control limits from Xbar-chart.  Dashed lines: control limits from I'-chart.

(ref:rpchart) P-chart of 24 random data from a binomial distribution (p = 0.1, subgroup size ranging from 100 to 120, seed = 1). Grey background: control limits from P-chart.  Dashed lines: control limits from I'-chart.

(ref:rppchart) P'-chart of 24 random data from a binomial distribution (p = 0.1, subgroup size ranging from 100 to 120, seed = 1). Grey background: control limits from P'-chart.  Dashed lines: control limits from I'-chart.

(ref:ruchart) U-chart from 24 random data from a poisson distribution (mean = 25, subgroup size ranging from 90 to 110, seed = 2). Grey background: control limits from U'-chart.  Dashed lines: control limits from I'-chart.

(ref:rupchart) U'-chart from 24 random data from a poisson distribution (mean = 25, subgroup size ranging from 90 to 110, seed = 2). Grey background: control limits from U'-chart.  Dashed lines: control limits from I'-chart.

<!-- Figure \@ref(fig:ichart): (ref:ichart) -->

<!-- Figure \@ref(fig:ipchart): (ref:ipchart) -->

<!-- Figure \@ref(fig:pchart): (ref:pchart) -->

<!-- Figure \@ref(fig:ppchart): (ref:ppchart) -->

<!-- Figure \@ref(fig:upchart): (ref:upchart) -->

<!-- Figure \@ref(fig:richart): (ref:richart) -->

<!-- Figure \@ref(fig:ripchart): (ref:ripchart) -->

<!-- Figure \@ref(fig:rxbarchart1): (ref:rxbarchart1) -->

<!-- Figure \@ref(fig:rxbarchart2): (ref:rxbarchart2) -->

<!-- Figure \@ref(fig:rxbarchart3): (ref:rxbarchart3) -->

<!-- Figure \@ref(fig:rpchart): (ref:rpchart) -->

<!-- Figure \@ref(fig:rppchart): (ref:rppchart) -->

<!-- Figure \@ref(fig:ruchart): (ref:ruchart) -->

<!-- Figure \@ref(fig:rupchart): (ref:rupchart) -->
